---
title: "Linear models"
author: "Instructor: Maxim Fedotov^[
    Credits: This handout is a re-adaptation of part of the materials created by
    Jordi Llorens, who taught this brush-up in previous years, themselves based on
    earlier materials by Hrvoje Stojic.
    Following the original materials, this instructional material is made available
    under the
    [Creative Commons Attribution 4.0 license](https://creativecommons.org/licenses/by/4.0/),
    and the accompanying code under the
    [MIT license](https://opensource.org/licenses/MIT).]"
date: "September 2026"
output:
  html_document: default
  pdf_document: default
---

```{r, knitr_options, include=FALSE}
    # loading in required packages
    if (!require("mvtnorm")) {
        install.packages("mvtnorm")
        library(mvtnorm)
    }
    if (!require("ggplot2")) {
        install.packages("ggplot2") 
        library(ggplot2)
    }
```


# Linear regression

Linear regression is a fundamental foundation of statistics. Despite its apparent simplicity, it is still used widely and is the basis for more sophisticated models. 

Consider a setting where you have a dataset with $n$ observations of a form $\{(y_i, \mathrm{x}_i)\}_{i=1}^{n}$ where $y_i$ is an output variable. Linear regression assumes linear dependence between the output $y$ and the vector of regressors $\mathrm{x}$ together with persistence of random noise. That is, 
$$y_i = \mathrm{x}_i^T \beta + \varepsilon_i$$
where $\beta \in \mathbb{R}^{p}$ is a vector of coefficients, $\varepsilon = (\varepsilon_1, \ldots, \varepsilon_n) \sim \mathbb{P}_\varepsilon$ is a noise vector.

The formulation can be rewritten in matrix notation:
$$y = X \beta + \varepsilon,$$
where $y \in \mathbb{R}^{n}$, $X \in \mathbb{R}^{n \times p}$ is a design matrix, $\varepsilon \in \mathbb{R}^{n}$.

# Simulating data

Let's first generate some artificial data for which we know the ground truth. We are going to write a function which allows us to create a dataset of arbitrarily many dimensions (i.e. a vector of coefficients can be of any length). We assume that the regressors come from multivariate normal distribution.

```{r simulate_lm}
simulate_lm <- function(n, b, noise_sd, rho_x = 0) {
    # How many regressors?
    p <- length(b)

    # First, define a covariance matrix (Sigma) for the regressors. 
    # Note: assume that an intercept is also included in b.
    Sigma <- diag(p - 1)
    Sigma[upper.tri(Sigma)] <- Sigma[lower.tri(Sigma)] <- rho_x
  
    # Draw the design matrix from a multivariate normal.
    x <- rmvnorm(n = n, sigma = Sigma)

    # Then generate y as a function of x and gaussian noise.
    # A linear model of the form: y = Xb + e;
    # use matrix multiplication operator    
    y <- cbind(1, x) %*% b + rnorm(n, mean = 0, sd = noise_sd)
    
    # put x and y in a dataframe
    data <- data.frame(y, x)

    # rename all the x variables so that they have the following format:
    # x1, x2, ...
    names(data) <- c("y", paste0("x", 1:(p - 1)))

    return(data)
}

# lets try it out
set.seed(1234)
data <- simulate_lm(200, c(5, 3, -2), 2, 0.3)
head(data)
```

# Illustrate the data 

We will first generate the data of lower dimension that can be easily visualized. We will use this data for the rest of the exercise.

```{r data_train}
set.seed(1234)
data_train <- simulate_lm(200, c(5, 3), 2)
head(data_train)
```

Use the `ggplot2` package to create a nicely formatted scatter plot of the data. 

```{r scatterplot, warning=F}
figure <- ggplot(data_train, aes(x1, y)) +
    geom_point(shape = 19, colour = "gray") +
    geom_smooth(method = lm, formula = y ~ x, se = TRUE) +
    theme_bw()

print(figure)
```


# Fitting linear models

The classic estimator of the linear regression model is the ordinary least squares estimator, defined as the minimizer of the residual sum of squares

$$\hat{\beta} = \textrm{argmin}_{\beta} (y - X\beta)^T (y - X\beta).$$

The estimator has a closed form solution whenever $X^TX$ is invertible.

$$\hat{\beta} = (X^T X)^{-1} X^T y.$$
It is worth mentioning that properties of this estimator depend on different conditions (e.g. see Gauss–Markov theorem).

You will use this estimator to compute the regression weights in your `lm_custom` function. It should produce the same coefficients as the `lm` function built-in R. But before that, use the `lm` function and fit a linear model to the data.

```{r}
# regress y on x
lmfit <- lm(y ~ x1, data = data_train)

# use the "summary" command on results to get a more detailed overview of the
# fit, you will get additional info like standard errors
summary(lmfit)

# extract the coefficients with SE's, t-statistics and p-values
# note that this is a matrix so we can extract anything we like from here
print(summary(lmfit)$coef) 
```

Let's obtain diagnostic plots by using plot command on the output of the
lm function (again, example of overloading the functions)
```{r diagnostic plots}
par(mfrow = c(2, 2))
plot(lmfit)
par(mfrow = c(1, 1))
```

Obtain predictions for the training data based on the fitted model
```{r}
pred_train <- fitted(lmfit)
head(pred_train)
```

and compute the mean square error between predictions and true values, y
```{r}
mse_train <- mean((pred_train - data_train$y)^2)
mse_train
```
How do you know how good is this? Compute now a mean square error 
for the base model - a simple mean of the observed y values
```{r}
mse_base <- mean((mean(data_train$y) - data_train$y)^2)
mse_base
```

so the model does a bit better obviously, however the true indicator of
how well the model does is the generalization performance, predictions
on the data it has not been fitted to. So, lets now create some new,
test data.
```{r}
set.seed(4321)
data_test <- simulate_lm(100, c(5, 3), 2)
head(data_test)
```
and verify how our model predicts on it
```{r}
pred_test <- predict(lmfit, data_test)
head(pred_test)
# compute the mean square error between predictions and true values, y
mse_test <- mean((pred_test - data_test$y)^2)
mse_test
# and benchmark against mse of just the sample mean of the training set
mse_base_test <- mean((mean(data_train$y) - data_test$y)^2 )
mse_base_test
```


Extra: next, fit the linear model on the data_train 
this time don't include the intercept in the model
and include an additional variable that is a square root of absolute value of X.
```{r}
lmfit2 <-lm(y ~ x1 + sqrt(abs(x1)) - 1, data = data_train)
lmfit2
summary(lmfit2)
```

## Custom function to compute coefficients of a linear model (OLS).

Now we will define our own function for fitting linear models, `lm_custom`, which computes the weights from the least squares solution above.

A note on how we compute it. We will not be inverting the $X^T X$ matrix. Instead, we will solve the linear system

$$(X^T X)\, \hat{\beta} = X^T y,$$

known as the *normal equations*. Here, this produces the same result as the analytic solution we discussed earlier in the handout. On badly conditioned problems the difference is not cosmetic though. Solving a system directly is both faster and numerically more stable than forming the inverse and then multiplying by it, so in R we write `solve(A, b)` rather than `solve(A) %*% b`.
```{r lm_custom}
lm_custom <- function(data, intercept = TRUE) {

    # we will assume that first column is the response variable
    # and all the others have the form x1, x2, ...
    
    # number of features
    m <- ncol(data) - 1

    # first we need to add a vector of 1's to our x (intercept),
    # if instructed by the intercept argument
    if (intercept) {
        X <- cbind(1, as.matrix(data[, 2:(m + 1)]))
    } else {
        X <- as.matrix(data[, 2:(m + 1)])
    }
    y <- data$y

    # Solve the normal equations. Note solve(A, b) rather than
    # solve(A) %*% b - same answer, better numerics.
    bhat <- solve(t(X) %*% X, t(X) %*% y)

    # compute the predictions for the training data, i.e. fitted values
    yhat <- X %*% bhat

    # compute the mean square error for the training data, between y and yhat
    mse <- mean((yhat - y)^2)

    return(list(coef = bhat, predictions = yhat, mse = mse))
}

# check out the function
lm_custom_fit <- lm_custom(data_train)
lm_custom_fit$coef

# compare it to the output of the lm function
coefficients(lmfit) 
```

# Illustrate the data and your model predictions

We will now create a plot where we illustrate our predictions.

```{r scatterplot_fit}

# first create an additional data frame with x1 variable from data_train as one
# column and predictions from the model, yhat, as a second column
data_predicted <- data.frame(x1 = data_train$x1, yhat = lm_custom_fit$predictions) 

# generate the figure
figure <- 
    ggplot(data_train, aes(y = y, x = x1)) + 
    geom_point(size = 1.5, color = "#992121") +

    # you will need to use geom_line, but now with
    # data_predicted, to illustrate the lm_custom fit
    geom_line(data = data_predicted, aes(x = x1, y = yhat), 
              color = "blue") +

    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks = element_line(
            linetype = 1,
            colour = "black", linewidth = 0.3
        ), 
        axis.text = element_text(colour = "black"),
        axis.text.x = element_text(vjust = 0.5)
    )

# show the figure in the report
print(figure)
```
